---
title: "openwashdata Newsletter - Issue 23"
date: "2026-09-23"
format: markdown
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
#| echo: false
#| include: false

new_pkgs <- c("solidwastekampala")

for (pkg in new_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pak(paste0("openwashdata/", pkg))
  }
}
```

```{=markdown}
[View this email in your browser]({{ email_url }})
```

Welcome back to our newsletter! This month, openwashdata has a new
look, and each issue now shows one dataset of the month with a chart
made from its data. The first is solidwastekampala, a new package on
household waste in Kampala. We also pass on two reflections by
Emmanuel Mhango on his year as Data Steward at BASEflow Malawi, and
thank Charles Niwagaba as our contributor of the month.

## 🎨 A new look for openwashdata

openwashdata has a new look. The website and this newsletter now share
the same colours, fonts and logos, in a light and a dark mode. The
landing page shows the dataset of the month, the latest blog posts and
a button to subscribe to this newsletter.

Every blog post is now also available as a PDF for download, so you can
read it offline or pass it on. Reports about openwashdata datasets can
carry the same look through our
[quarto-owd extension](https://github.com/openwashdata/quarto-owd),
which produces PDF and Word documents.

[See the new website](https://openwashdata.org) and
[browse the brand guide](https://openwashdata.github.io/brand/)

## 📊 Dataset of the Month: solidwastekampala

Our first dataset of the month is solidwastekampala, a new data
package from Makerere University with the quantity and
composition of domestic solid waste from 103 households in three
Kampala parishes of low, middle and high income, weighed over one
week. Waste per person rises with income, from 0.43 to 0.98 kg per
day, and food makes up at least three quarters of the mass in every
parish.

The package holds the raw data as collected, before the analysis in
the paper the team published this month in Frontiers in
Environmental Science, [Quantity and composition of domestic solid
waste in Kampala City as influenced by socioeconomic
factors](https://doi.org/10.3389/fenvs.2026.1889921). Anyone can
start from the same records the authors worked with.

- `solidwastekampala` -
  <https://openwashdata.github.io/solidwastekampala>:
  `r packageDescription("solidwastekampala")$Title`

```{r}
#| label: card-solidwastekampala
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
source("R/card.R")

card <- owd_card_stats("solidwastekampala", countries = "Kampala, Uganda")

groups <- c(
  waste_food_kg = "Food",
  waste_garden_kg = "Garden and wood",
  waste_wood_kg = "Garden and wood",
  waste_polythene_kg = "Plastics and polythene",
  waste_plastics_kg = "Plastics and polythene",
  waste_paper_kg = "Paper and textiles",
  waste_textiles_kg = "Paper and textiles",
  waste_glass_kg = "Glass and metals",
  waste_metals_kg = "Glass and metals",
  waste_other_kg = "Other"
)

shares <- card$data |>
  select(income_level, parish, all_of(names(groups))) |>
  pivot_longer(starts_with("waste_"), names_to = "category",
    values_to = "kg") |>
  mutate(group = unname(groups[category])) |>
  summarise(kg = sum(kg, na.rm = TRUE),
    .by = c(income_level, parish, group)) |>
  mutate(share = kg / sum(kg), .by = income_level)

group_order <- shares |>
  summarise(kg = sum(kg), .by = group) |>
  arrange(desc(kg)) |>
  pull(group)

parish_labels <- shares |>
  distinct(income_level, parish) |>
  arrange(income_level) |>
  mutate(label = paste0(tools::toTitleCase(as.character(income_level)),
    " income\n(", parish, ")"))

shares <- shares |>
  left_join(parish_labels, by = c("income_level", "parish")) |>
  mutate(
    group = factor(group, levels = rev(group_order)),
    label = factor(label, levels = parish_labels$label)
  )

# chart-1 to chart-6 from openwashdata/brand
chart_colours <- c("#8a3a8a", "#b04208", "#CF9E40", "#006840", "#FD7AAD",
  "#304FC2")

p <- ggplot(shares, aes(x = label, y = share, fill = group)) +
  geom_col(width = 0.62) +
  scale_y_continuous(labels = scales::percent,
    expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = chart_colours, breaks = group_order) +
  labs(
    title = "Food makes up most of the household waste in Kampala",
    subtitle = paste("Share of waste mass by category,",
      "103 households in three parishes, one week"),
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_owd() +
  theme(legend.position = "right", legend.key.size = unit(0.9, "lines"))

img <- owd_card_save(p, card, alt = "Stacked bar chart of the share of
  household waste mass by category for the low, middle and high income
  parishes in Kampala. Food makes up 75 to 82 percent in each parish.
  Garden and wood, plastics and polythene, other waste, paper and
  textiles, and glass and metals share the rest.")
```

`r img`

`r owd_card_line(card)`

The package took a couple of hours to build. Our data publishing
pipeline has matured to the point where we can take a dataset in any
format, with minimal documentation from the author, and turn it into
a complete data package with documented variables, a website and a
citation. This is possible because agentic AI tools now work together
with the data publishing software we have invested in for years.
Going forward, openwashdata will focus more on giving others the
opportunity to benefit from these possibilities.

## 📄 Blog Highlights: a year as Data Steward at BASEflow Malawi

Emmanuel Mhango works as Data Steward at BASEflow Malawi within the
openwashdata project and has published five WASH datasets with us. He
wrote two reflections on that year, first published on LinkedIn and
now on our blog.

The first post starts with one question, what data does the
organization actually have, and follows the answer from a data
inventory to a Data Management Strategy, and to the lesson that data
stewardship begins with people. The second looks at what happened once
BASEflow started publishing data: the questions Open Science raised
about ownership, permission and documentation, and a groundwater
dataset that UNICEF Malawi used to identify where boreholes need
attention.

- [I thought my job was about data. It turned out to be about
  people.](https://openwashdata.org/pages/blog/posts/2026-08-11-job-about-data/)
- [Beyond publishing data: what we learned building an open science
  culture at BASEflow
  Malawi](https://openwashdata.org/pages/blog/posts/2026-08-18-beyond-publishing-data/)

## 🌟 Contributor of the Month

Our contributor of the month is Charles Niwagaba of Makerere
University, who shared the household solid waste data behind the
solidwastekampala package with the community. The package names the
whole team that collected the data in Kampala as authors. Thank you,
Charles.

Charles and his co-authors published their raw data with us before
the analysis, and their paper came out in Frontiers in Environmental
Science on 11 September. This is the second publication where
authors have worked with us to publish the raw data behind a paper.
We are looking for more volunteers. If you have a manuscript in
progress and data you can share, [share your data with
us](https://openwashdata.org/pages/blog/posts/2024-05-17-data-publication-1/).

## Get Involved

We believe the openwashdata project prospers when we have **YOU**
work together and promote open science and data practice! No matter
what background you are from, we come up with some ways for you to
get involved:

- [Join our chatroom to meet
  people!](https://openwashdata.org/pages/get-started/chat/)
- [Share your WASH data with
  us](https://openwashdata.org/pages/blog/posts/2024-05-17-data-publication-1/)
- Spread the word. Forward this email.
- Got more ideas? [Leave us a message on Matrix to
  collaborate!](https://matrix.to/#/%23openwashdata-lobby:staffchat.ethz.ch)
