---
title: "openwashdata Newsletter - Issue 21"
date: "2026-03-06"
format: markdown
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
#| echo: false
#| include: false

new_pkgs <- c("mwgroundwaterdata", 
              "mwefloodresponse",
              "waterpointfunctionality")

devtools::install_github(paste0("openwashdata", "/", new_pkgs))

```

Welcome to the first edition of 2026 of our newsletter! This month, we present the updated GLAAS package, introduce our new R package for data sharing, share an update in our ongoing collaboration with BASEFlow Malawi, and highlight an opportunity to join our team through an internship.

## 🔍 Updated GLAAS package

Thanks to the work of [Colin](https://openwashdata.org/about/colin/) and [Nicolò](https://openwashdata.org/about/nicolo/) we now have an updated version of the GLAAS R package. It provides comprehensive access to the UN-Water Global Analysis and Assessment of Sanitation and Drinking-water (GLAAS) dataset. The WHO GLAAS survey collects data on water, sanitation, and hygiene (WASH) systems, policies, and financing from countries worldwide.

Visit the package's website for more information and some insightful visualizations: <https://openwashdata.github.io/glaas/>

## 📦 Blog Highlight: Data Sharing Package

Our latest blog post by [Adriana](https://openwashdata.org/about/adriana/) introduces fairenough: our brand new R package for data publication and sharing. It also goes through the motivation for developing our CRAN package washr and this next revamped iteration, which minimizes users' input and leverages LLMs to automatically produce data dictionaries.

The blog post includes more details about the most important features and covers our presentation of this package at LatinR 2025 last year.

Explore fairenough: <https://openwashdata.github.io/fairenough/>

[Read the full blog post here](https://openwashdata.org/pages/blog/posts/2026-02-10-fairenough/)

## 🤝 Continued partnership with BASEFlow Malawi

Over the past year, [Emmanuel](https://openwashdata.org/about/emmanuel/) has led data quality and publication efforts for WASH datasets and supported open data workflows in the global WASH sector as part of BASEFlow Malawi. Through our partnership with BASEFlow, he has contributed over 15 high-quality datasets to openwashdata. We're pleased to announce that this collaboration will continue for another year!

Some of the latest data packages contributed by Emmanuel to the open WASH data collection include:

- `mwgroundwaterdata` - <https://openwashdata.github.io/mwgroundwaterdata>: `r packageDescription(new_pkgs[[1]])$Title`
- `mwefloodresponse` - <https://openwashdata.github.io/mwefloodresponse>: `r packageDescription(new_pkgs[[2]])$Title`
- `waterpointfunctionality` - <https://openwashdata.github.io/waterpointfunctionality>: `r packageDescription(new_pkgs[[3]])$Title`

[You can see our published datasets here](https://openwashdata.org/pages/gallery/data/)

## 🌟 Contributor of the Month

This month we want to recognize **Nicolò Massari**, who finished his internship with us at the end of 2025. Nicolò designed and implemented the core features of fairenough and we're excited to continue the development of the package!

## 💼 Join Our Team: Data Science / Software Engineer Internship

We're looking for a new intern to join our Open Science team at Global Health Engineering, ETH Zurich! In this role (80–100%), you'll support data cleaning and extraction processes, contribute to openwashdata packages including washr and fairenough, and help compile metadata for ETH's Open Research Data program.

We're looking for someone who cares about concise, reusable code, is comfortable with data science tools like Git, GitHub, and the R Tidyverse, and stays up to date on the latest developments in LLMs. The preferred start date is July 1, 2026, and applications are due by April 30, 2026.

Questions about the position can be directed to [Adriana](https://openwashdata.org/about/adriana/).

[View the full posting and apply here](https://jobs.ethz.ch/job/view/JOPG_ethz_AImLu4H3zIaJUVZuBm)

## Get Involved

We believe the openwashdata project prospers when we have **YOU** work together and promote open science and data practice! No matter what background you are from, we come up with some ways for you to get involved:

- [Join our chatroom to meet people!](https://openwashdata.org/pages/get-started/chat/)
- [Share your WASH data with us](https://openwashdata.org/pages/blog/posts/2024-05-17-data-publication-1/)
- Spread the word. Forward this email. 
- Got more ideas? [Leave us a message on Matrix to collaborate!](https://matrix.to/#/%23openwashdata-lobby:staffchat.ethz.ch)