---
title: "openwashdata Newsletter - Issue 24"
date: "2026-10-21"
format:
  markdown:
    wrap: none
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
#| echo: false
#| include: false

new_pkgs <- c("gwinfraassess")

for (pkg in new_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    pak::pak(paste0("openwashdata/", pkg))
  }
}
```

```{=markdown}
[View this email in your browser]({{ email_url }})
```

Welcome back to our newsletter! This month's dataset of the month
comes from BASEflow Malawi and looks at the state of 123 groundwater
monitoring wells. Our blog tells how the solidwastekampala package went
from a raw spreadsheet to a published data package in a day, and how
we used AI along the way. We also thank Alex Y. Katukiza as our
contributor of the month.

## 📊 Dataset of the Month: gwinfraassess

gwinfraassess holds an assessment of 123 groundwater monitoring wells in
Malawi, carried out by BASEflow Malawi in 2022 and 2023 and published by
[Emmanuel Mhango](https://openwashdata.org/about/emmanuel/). For each well,
field teams recorded the depth and the water level, the state of the
automatic water level logger, the measuring equipment, how the
community relates to the well, and what the well needs to keep working.

Every well but one has an improvement plan. The action named most often
is a trained person to read the gauge, in 76 of 122 wells, followed by
a manual dipper in 74 wells. Many wells have a logger, but on the day of
the assessment the data could be downloaded from only 10 of 67 loggers.

- `gwinfraassess` - <https://openwashdata.github.io/gwinfraassess>:
  `r packageDescription("gwinfraassess")$Title`

```{r}
#| label: card-gwinfraassess
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
source("R/card.R")

card <- owd_card_stats("gwinfraassess", countries = "Malawi")

# The same chart as the dataset of the month on openwashdata.org
# (index.qmd, Observable Plot), drawn once as a static image: the share
# of wells with an improvement plan that need each action.
action_groups <- c(
  "Identify and training gauge reader" = "People",
  "Provide manual dipper" = "Equipment",
  "Replacement of the level logger" = "Equipment",
  "Installation of a new logger" = "Equipment",
  "Replace barologger" = "Equipment",
  "Rehabilitation of the civil works" = "Structure",
  "Replace end cap" = "Structure",
  "Replacement of the cover and hinges" = "Structure",
  "Welding on the existing hinges" = "Structure",
  "Replace and weld the hinges" = "Structure",
  "Redevelop the monitoring well" = "Structure",
  "Replace the well" = "Replace or retire",
  "Discard monitoring well" = "Replace or retire",
  "Discard as monitoring well and donate to the community" =
    "Replace or retire"
)

action_labels <- c(
  "Identify and training gauge reader" = "Find and train a gauge reader",
  "Provide manual dipper" = "Provide a manual dipper",
  "Replacement of the level logger" = "Replace the level logger",
  "Installation of a new logger" = "Install a new logger",
  "Replace barologger" = "Replace the barologger",
  "Rehabilitation of the civil works" = "Repair the civil works",
  "Replace end cap" = "Replace the end cap",
  "Replacement of the cover and hinges" = "Replace cover and hinges",
  "Welding on the existing hinges" = "Weld the existing hinges",
  "Replace and weld the hinges" = "Replace and weld the hinges",
  "Redevelop the monitoring well" = "Redevelop the well",
  "Replace the well" = "Replace the well",
  "Discard monitoring well" = "Discard the well",
  "Discard as monitoring well and donate to the community" =
    "Discard, give to community"
)

wells <- card$data |>
  mutate(well = row_number()) |>
  filter(!is.na(improvement_plan))

n_plans <- nrow(wells)

actions <- wells |>
  separate_longer_delim(improvement_plan, ",") |>
  mutate(action = gsub("\\s+", " ", trimws(improvement_plan))) |>
  filter(action %in% names(action_groups)) |>
  distinct(well, action) |>
  count(action, name = "n") |>
  mutate(
    group = factor(unname(action_groups[action]),
      levels = c("People", "Equipment", "Structure", "Replace or retire")),
    label = unname(action_labels[action]),
    share = round(100 * n / n_plans)
  ) |>
  arrange(desc(n)) |>
  mutate(label = factor(label, levels = rev(label)))

# chart-1, chart-3, chart-2 and a neutral grey from openwashdata/brand
palette <- c(
  "People" = "#8a3a8a", "Equipment" = "#CF9E40",
  "Structure" = "#b04208", "Replace or retire" = "#b9b9b9"
)
font <- "Atkinson Hyperlegible Next"

p <- ggplot(actions, aes(x = share, y = label, fill = group)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = paste0(share, "%")), hjust = 0, nudge_x = 0.8,
    family = font, size = 3.2, colour = owd_colors[["dark"]]) +
  scale_fill_manual(values = palette, name = NULL) +
  scale_x_continuous(limits = c(0, 70), breaks = seq(0, 60, 20),
    expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "The most common fix is a person",
    subtitle = paste("Share of the", n_plans, "wells with an improvement",
      "plan that need each action"),
    x = "% of wells", y = NULL
  ) +
  theme_owd() +
  theme(
    text = element_text(family = font),
    panel.grid.major.x = element_line(colour = "#e3e3e3"),
    panel.grid.major.y = element_blank(),
    legend.position = "top",
    legend.justification = "left",
    axis.title.x = element_text(colour = owd_colors[["gray"]], size = 10)
  )

img <- owd_card_save(p, card, height = 5.2, alt = "Bar chart of the
  improvement actions recommended for 122 groundwater monitoring wells
  in Malawi. 62 percent need a trained gauge reader, 61 percent a manual
  dipper, 46 percent a replaced level logger and 42 percent repaired
  civil works.")
```

`r img`

`r owd_card_line(card)`

### Our question for you

In the Malawi wells, the most common need was a person trained to read
the gauge. We would like to know how this looks in your work. Choose one
answer with a click, and we will share the results in the next issue.

**What is the biggest gap in the data you work with?**

```{=markdown}
{{ survey.monitoringgap }}
```

## 📄 Blog Highlights: from a raw spreadsheet to a published data package

In July, the household waste data from Kampala that we featured last
month arrived as an Excel workbook with merged headers, summary rows and
the typos of manual data entry. A day later it was a reviewed and
citable R data package with a DOI. The new post retells that day from
the record the repository keeps, including every prompt we sent to the
AI. Data cleaning was the smallest part of the work, and the post shows
where the rest went.

The post ends with a statement of how we used AI for the package and for
the post itself. We want to make such statements a standard part of our
work.

- [From a raw spreadsheet to a published data package in a
  day](https://openwashdata.org/pages/blog/posts/2026-10-20-solidwastekampala-story/)

## 🌟 Contributor of the Month

Our contributor of the month is [Alex Y.
Katukiza](https://orcid.org/0000-0003-1774-8237) of Makerere University,
the lead author of the household waste study behind solidwastekampala.
While we prepared the package, Alex confirmed the ORCID iDs of his
co-authors and settled the spelling of their names, so that every author
is credited correctly in the citation. Our blog post this month tells
the story of that package. Thank you, Alex.

## Get Involved

We believe the openwashdata project prospers when we have **YOU**
work together and promote open science and data practice! No matter
what background you are from, we come up with some ways for you to
get involved:

- [Join our chatroom to meet
  people!](https://openwashdata.org/pages/get-started/chat/)
- [Share your WASH data with
  us](https://openwashdata.org/pages/blog/posts/2024-05-17-data-publication-1/)
- Spread the word. Forward this email.
- Got more ideas? [Leave us a message on Matrix to
  collaborate!](https://matrix.to/#/%23openwashdata-lobby:staffchat.ethz.ch)
